---
title: "DANL 310 Final Project: What Predicts NFL Success?"
author: "Ryan Horn"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: show
execute:
  warning: false
  message: false
---

## 1. Introduction

NFL teams can win in different ways, but not all winning styles are equally sustainable.
This project asks a simple question:

**Are playoff teams mainly offense-led, defense-led, or balanced on both sides of the ball?**

We use season-level NFL team records from the nflverse data repository to build a clear story around team performance, style, and postseason outcomes.

## 2. Data

The dataset comes from nflverse standings data:

- Source: <https://github.com/nflverse/nfldata/raw/master/data/standings.csv>
- Unit of analysis: one team in one season
- Core fields: wins, losses, ties, points scored, points allowed, point differential, and playoff indicator
- Time window: 2002 through the latest completed season in the file

```{r}
library(tidyverse)
library(scales)

source("R/nfl_data_prep.R")

standings_raw <- load_nfl_standings()
team_season <- standings_raw |>
  prep_nfl_team_season(min_season = 2002) |>
  add_team_style()

story <- build_story_data(team_season)
latest_season <- max(team_season$season)

team_season |>
  summarise(
    min_season = min(season),
    max_season = max(season),
    team_seasons = n(),
    unique_teams = n_distinct(team),
    playoff_share = mean(playoff_flag)
  )
```

## 3. Data Transformation and Descriptive Statistics

### Transformations used

- Filtered to completed team-seasons (`games_played >= 16`)
- Created per-game metrics for fair comparisons across 16-game and 17-game eras:
  - `points_for_pg`
  - `points_allowed_pg`
  - `point_diff_pg`
- Created a `playoff_flag` indicator
- Built team style categories by comparing each team to league averages in the same season:
  - `Balanced contender`
  - `Offense-led`
  - `Defense-led`
  - `Needs both sides`

```{r}
team_season |>
  count(style, sort = TRUE)
```

A quick descriptive check: point differential and winning percentage are strongly related.

```{r}
team_season |>
  summarise(correlation = cor(point_diff_pg, win_pct, use = "complete.obs"))
```

## 4. Data Visualization and Storytelling

### Figure 1. Scoring trend in the NFL

```{r}
ggplot(story$league_trend, aes(x = season, y = avg_points_for_pg)) +
  geom_line(color = "steelblue", linewidth = 1) +
  geom_smooth(method = "loess", se = FALSE, color = "darkorange", linewidth = 1) +
  labs(
    title = "League scoring has generally risen in the modern era",
    subtitle = "Average points scored per team per game",
    x = "Season",
    y = "Points per game"
  ) +
  scale_y_continuous(labels = number_format(accuracy = 0.1)) +
  theme_minimal(base_size = 12)
```

Scoring environments have changed across eras, which is why per-game metrics and season-relative style definitions are useful.

### Figure 2. Point differential is the backbone of winning

```{r}
ggplot(team_season, aes(x = point_diff_pg, y = win_pct, color = playoff_flag)) +
  geom_point(alpha = 0.7) +
  geom_smooth(method = "lm", se = FALSE, color = "gray30", linewidth = 1) +
  scale_color_manual(values = c("TRUE" = "#1b9e77", "FALSE" = "#d95f02"), labels = c("No", "Yes")) +
  scale_y_continuous(labels = percent_format(accuracy = 1)) +
  labs(
    title = "Teams with stronger point differential win more",
    subtitle = "Playoff teams cluster in the high differential / high win-rate region",
    x = "Point differential per game",
    y = "Winning percentage",
    color = "Made playoffs"
  ) +
  theme_minimal(base_size = 12)
```

This plot shows that total team quality (scoring more while allowing less) is tightly connected to regular-season success.

### Figure 3. Which team style makes the playoffs most often?

```{r}
ggplot(story$playoff_style, aes(x = style, y = playoff_rate, fill = era)) +
  geom_col(position = position_dodge(width = 0.8), width = 0.7) +
  scale_y_continuous(labels = percent_format(accuracy = 1), limits = c(0, 1)) +
  labs(
    title = "Balanced teams have the highest playoff rates",
    subtitle = "Playoff rate by style and era",
    x = "Team style",
    y = "Playoff rate",
    fill = "Era"
  ) +
  theme_minimal(base_size = 12) +
  theme(axis.text.x = element_text(angle = 15, hjust = 1))
```

The most consistent pattern is that **balanced contenders** are the safest path to playoff football.
Offense-led and defense-led teams can still succeed, but they are typically less reliable than teams that are above league average on both sides.

### Figure 4. Recent outliers (latest season)

```{r}
team_season |>
  filter(season == latest_season) |>
  ggplot(aes(x = points_for_pg, y = points_allowed_pg, label = team, color = playoff_flag)) +
  geom_point(size = 3, alpha = 0.8) +
  geom_text(size = 3, check_overlap = TRUE, vjust = -0.8) +
  scale_color_manual(values = c("TRUE" = "#1b9e77", "FALSE" = "#d95f02"), labels = c("No", "Yes")) +
  labs(
    title = paste("Style map for", latest_season),
    subtitle = "Right is better offense; lower is better defense",
    x = "Points scored per game",
    y = "Points allowed per game",
    color = "Made playoffs"
  ) +
  theme_minimal(base_size = 12)
```

This final view highlights team archetypes in the most recent season and helps connect the full-league trend to concrete team examples.

## 5. Conclusion

Across modern NFL seasons, one message is clear:

- Point differential strongly tracks winning.
- Playoff teams are disproportionately balanced teams.
- One-sided strengths (only offense or only defense) can work, but they are less dependable.

So the best storytelling takeaway is not "offense always wins" or "defense always wins".
Instead, **balance wins most consistently**.

## 6. References (Optional)

- nflverse `standings.csv`: <https://github.com/nflverse/nfldata/raw/master/data/standings.csv>
