---
title: DANL Project
subtitle: "Data-Driven Mastery: Road to excellence"
author: 
  - Ryan Horn
  - Dylan Emler
  - Ethan Hagan
  - Rachel Rabkin
  - Kara Nash

toc: true
toc-title: "Table of Contents"
toc-depth: 2
number-sections: true

fig-width: 9

execute:
  code-tools: true
  message: false
  warning: false

from: markdown+emoji
---

```{r setup}
#| include: false
library(tidyverse)
library(skimr)
library(ggthemes)
library(hrbrthemes)

theme_set(theme_ipsum()+
          theme(strip.background =element_rect(fill="lightgray"),
                axis.title.x = 
                  element_text(angle = 0,
                               size = rel(1.33),
                               margin = margin(10,0,0,0)),
                axis.title.y = 
                  element_text(angle = 0,
                               size = rel(1.33),
                               margin = margin(0,10,0,0))
                )
          )
```

## Introduction

About this project :clap: 

  In this project my group and I are using a data set of motor cars to determine the fastest cars, then we will decipher some parameters by what makes them the best.

# Data

```{r}
data("mtcars")
```


The data was extracted from the 1974 Motor Trend US magazine, and comprises fuel consumption and 10 aspects of automobile design and performance for 32 automobiles (1973–74 models).

## Summary Statistics
```{r}
rmarkdown::paged_table(mtcars)
```
### Summary Statistics
```{r}
skim(mtcars) %>% 
  select(-n_missing)
```
# Visualization #1

```{r}
car_models <- rownames(mtcars)

ggplot(mtcars, aes(x = reorder(car_models, -qsec), y = qsec)) +
  geom_point(color = "blue") +
  labs(title = "1/4 Mile Times by Car Name",
       x = "Car Name",
       y = "1/4 Mile Time") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
car_models <- rownames(mtcars)
mtcars$cyl <- factor(mtcars$cyl)

```

## Explanation


This graph gives a visual representation of the quarter mile times of various motor cars in descending order. 


# Visualization #2
```{r}
ggplot(mtcars, aes(x = factor(cyl), y = qsec, fill = factor(cyl))) +
  geom_boxplot(alpha = 0.7) +
  geom_point(position = position_jitter(width = 0.3), color = "black", size = 2) +
  labs(title = "1/4 Mile Time by Number of Cylinders",
       x = "Number of Cylinders",
       y = "1/4 Mile Time",
       fill = "Number of Cylinders") +
  scale_fill_discrete(name = "Number of Cylinders") +
  theme_minimal()

```
## Explanation


This graph allows you to visually see the difference in quarter mile time and how it varies by the number of cylinders in the vehicle. As you can see the vehicles with fewer cylinders can generally be seen with a slower quarter mile time.


# Visualization #3
```{r}
ggplot(mtcars, aes(x = wt, y = qsec, color = cyl)) +
  geom_point() +
  labs(title = "Scatter Plot of Quarter Mile Time vs. Weight by Number of Cylinders", 
       x = "Weight (1000 lbs)",
       y = "Quarter Mile Time") +
  scale_color_discrete(name = "Number of Cylinders")

```
## Explanation


This visualization is similar to the last one except that this graph also shows the weight distribution among the vehicles, generally we can see that cars with more cylinders will be heavier than their counterparts.


# Visualization #4
```{r}
ggplot(mtcars, aes(x = hp, y = qsec, color = factor(vs))) +
  geom_point(size = 3) +
  labs(title = "Quarter Mile Time by Horsepower and Engine Type",
       x = "Horsepower",
       y = "Quarter Mile Time") +
  scale_color_discrete(name = "Engine Type", labels = c("V-shaped", "Straight")) +
  theme_minimal()

```
## Explanation 


Here we can decipher the quarter mile time in terms of the vehicles horse power and engine type, as we can see, as horse power grows, the quarter mile time goes down. At the same time most vehicles contain a V-Shaped engine that how the faster quarter mile times.


# Visualization #5
```{r} 
library(dplyr)
library(ggplot2)

top_5_fastest <- mtcars %>%
  arrange(qsec) %>%
  head(5)

print(top_5_fastest[, c("mpg", "cyl", "hp", "wt", "qsec")])

top_5_fastest$car_names <- rownames(top_5_fastest)

variables_to_plot <- c("cyl", "hp", "wt")
plot_list <- lapply(variables_to_plot, function(var) {
  ggplot(top_5_fastest, aes(x = reorder(car_names, qsec), y = get(var), fill = factor(cyl))) +
    geom_bar(stat = "identity", position = "dodge") +
    labs(title = paste("Top 5 Fastest Cars -", var),
         x = "Car",
         y = var,
         fill = "Cylinders") +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
})

library(patchwork)
plot_list[[1]] + plot_list[[2]] + plot_list[[3]]

```
## Explanation

Finally, this visualization shows how 3 variables (weight, cylinders, and horsepower) differ in each fastest car (from left to right).


