---
title: "NFL Success Dashboard"
format: dashboard
execute:
  warning: false
  message: false
---

```{r}
#| context: setup
#| include: false
required_pkgs <- c("dplyr", "ggplot2", "readr", "scales")
missing_pkgs <- required_pkgs[!vapply(required_pkgs, requireNamespace, logical(1), quietly = TRUE)]

if (length(missing_pkgs) > 0) {
  stop(
    paste0(
      "Missing required package(s): ",
      paste(missing_pkgs, collapse = ", "),
      ". Install with: install.packages(c(",
      paste(sprintf('\"%s\"', missing_pkgs), collapse = ", "),
      "))"
    )
  )
}

library(dplyr)
library(ggplot2)
library(scales)

source("R/nfl_data_prep.R")

team_season <- load_nfl_standings() |>
  prep_nfl_team_season(min_season = 2002) |>
  add_team_style()

story <- build_story_data(team_season)
latest_season <- max(team_season$season)
latest_data <- team_season |>
  filter(season == latest_season)

best_team <- latest_data |>
  arrange(desc(point_diff_pg)) |>
  slice(1)
```

## {.sidebar width="300px"}

This dashboard gives a quick summary of the NFL storytelling project using season-level team data from nflverse.

- Coverage: 2002 through `r latest_season`
- Focus: scoring, point differential, team style, and playoff outcomes
- Full write-up: [NFL Story Page](nfl-story-project.html)

The separate Shiny app is reserved for the fully interactive version of this project.

## Row

### 

```{r}
#| content: valuebox
#| title: "Latest Season"
list(
  value = latest_season,
  icon = "calendar-event",
  color = "primary"
)
```

### 

```{r}
#| content: valuebox
#| title: "Playoff Teams"
list(
  value = sum(latest_data$playoff_flag),
  icon = "trophy",
  color = "success"
)
```

### 

```{r}
#| content: valuebox
#| title: "Best Point Differential"
list(
  value = paste0(best_team$team, " (", round(best_team$point_diff_pg, 1), ")"),
  icon = "graph-up-arrow",
  color = "info"
)
```

## Row

### Latest Season Style Map

```{r}
latest_data |>
  ggplot(aes(x = points_for_pg, y = points_allowed_pg, color = style, label = team)) +
  geom_point(size = 3, alpha = 0.9) +
  geom_text(check_overlap = TRUE, nudge_y = 0.25, size = 3) +
  labs(
    title = paste("Offense and defense profile in", latest_season),
    x = "Points scored per game",
    y = "Points allowed per game",
    color = "Team style"
  ) +
  theme_minimal(base_size = 12)
```

### Winning Signal Across Eras

```{r}
team_season |>
  ggplot(aes(x = point_diff_pg, y = win_pct, color = playoff_flag)) +
  geom_point(size = 2.5, alpha = 0.7) +
  geom_smooth(method = "lm", se = FALSE, color = "gray30") +
  scale_color_manual(
    values = c("TRUE" = "#1b9e77", "FALSE" = "#d95f02"),
    labels = c("No", "Yes")
  ) +
  scale_y_continuous(labels = percent_format(accuracy = 1)) +
  labs(
    title = "Point differential remains the strongest simple winning signal",
    x = "Point differential per game",
    y = "Winning percentage",
    color = "Made playoffs"
  ) +
  theme_minimal(base_size = 12)
```

## Row

### Playoff Rate by Team Style

```{r}
story$playoff_style |>
  ggplot(aes(x = style, y = playoff_rate, fill = era)) +
  geom_col(position = position_dodge(width = 0.8), width = 0.7) +
  scale_y_continuous(labels = percent_format(accuracy = 1), limits = c(0, 1)) +
  labs(
    title = "Balanced teams reach the playoffs more often",
    x = "Team style",
    y = "Playoff rate",
    fill = "Era"
  ) +
  theme_minimal(base_size = 12) +
  theme(axis.text.x = element_text(angle = 15, hjust = 1))
```

### Latest Season Leaders

```{r}
latest_data |>
  transmute(
    Team = team,
    Wins = wins,
    Losses = losses,
    `Win %` = percent(win_pct, accuracy = 0.1),
    `Pts/Game` = round(points_for_pg, 1),
    `Opp Pts/Game` = round(points_allowed_pg, 1),
    `Diff/Game` = round(point_diff_pg, 1),
    Playoffs = if_else(playoff_flag, "Yes", "No"),
    Style = style
  ) |>
  arrange(desc(Wins), desc(`Diff/Game`)) |>
  head(10)
```

## Row

### League Scoring Trend

```{r}
story$league_trend |>
  ggplot(aes(x = season, y = avg_points_for_pg)) +
  geom_line(color = "steelblue", linewidth = 1) +
  geom_smooth(method = "loess", se = FALSE, color = "darkorange", linewidth = 1) +
  labs(
    title = "League scoring has generally risen in the modern era",
    x = "Season",
    y = "Average points per game"
  ) +
  theme_minimal(base_size = 12)
```
